---
title: "Aligning landmark configurations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Aligning landmark configurations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  message = FALSE,
  warning = FALSE
)
```

```{r load}
library(gprocrustes)
```

You have the same landmarks recorded more than once: different sessions,
sites, or observers, each in its own pose. You want one shared shape
and a record of how each recording was moved onto it.

Rows are entities (landmarks). Columns are dimensions. The package
moves a configuration by the right action $T(X)=sXR+\mathbf{1}t$.

## Can I see a shared shape first?

Three noisy copies of an eight-landmark outline. One is rotated and
translated; one is also scaled.

```{r make-recordings}
set.seed(1)
shape <- matrix(
  c(0, 0, 1, 0.1, 1.4, 0.8, 1, 1.4, 0.2, 1.5, -0.2, 1, -0.3, 0.4, 0.2, -0.1),
  8, 2, byrow = TRUE
)
shape <- scale(shape, scale = FALSE)
rot <- function(th) matrix(c(cos(th), sin(th), -sin(th), cos(th)), 2, 2)
recordings <- list(
  site_a = shape + matrix(rnorm(16, sd = 0.03), 8, 2),
  site_b = shape %*% rot(0.8) + matrix(c(0.4, -0.2), 8, 2, byrow = TRUE) +
    matrix(rnorm(16, sd = 0.03), 8, 2),
  site_c = 1.35 * (shape %*% rot(-0.5)) + matrix(c(-0.3, 0.5), 8, 2, byrow = TRUE) +
    matrix(rnorm(16, sd = 0.03), 8, 2)
)
```

`gpa()` fits a consensus and a transform for each recording.

```{r fit-gpa}
fit <- gpa(recordings, transform = "similarity")
fit
```

```{r check-default-fit, include = FALSE}
stopifnot(is.finite(fit$objective), fit$objective > 0)
stopifnot(fit$solver == "gower_bcd")
stopifnot(fit$optimality_status == "blockwise_stationary")
stopifnot(identical(certify(fit)$status, "unavailable"))
stopifnot(is.finite(quotient_distance(consensus(fit), consensus(fit, gauge = "canonical"))))
```

The overlay is the first thing to look at: aligned recordings around
the consensus.

```{r plot-overlay}
plot(fit, type = "overlay")
```

`consensus()` is a representative of an equivalence class. Changing
the gauge does not change the residual energy.

```{r gauge-is-presentation}
native <- consensus(fit, gauge = "native")
canonical <- consensus(fit, gauge = "canonical")
c(
  objective = glance.gpa_fit(fit)$objective,
  after_canonicalize = glance.gpa_fit(canonicalize(fit))$objective,
  quotient = quotient_distance(native, canonical)
)
```

## What did the engine actually claim?

`gpa()` is a compiler, not a universal optimizer. This similarity
problem used Gower block-coordinate descent. The allowed claim is
blockwise stationarity, not a global certificate.

```{r explain-claim}
explain_solver(fit)
certify(fit)
```

`glance()` is the one-row summary of those existing diagnostics.
`tidy()` is one row per recording: scale and translation already
stored on the transforms.

```{r broom-tables}
glance.gpa_fit(fit)
tidy.gpa_fit(fit)
```

## What if one recording is a fixed reference?

Two views and an `anchor` use the exact pairwise polar kernel.

```{r pairwise-anchor}
pair <- gpa(
  recordings[c("site_a", "site_b")],
  transform = proc_similarity(group = "SO"),
  anchor = "site_a"
)
c(solver = pair$solver, optimality = pair$optimality_status)
```

```{r check-pairwise, include = FALSE}
stopifnot(identical(pair$solver, "pairwise_polar"))
stopifnot(identical(pair$optimality_status, "exact_closed_form"))
```

That closed form is also the standalone `procrustes()` kernel.

```{r pairwise-kernel}
procrustes(
  recordings$site_b,
  recordings$site_a,
  transform = proc_similarity(group = "SO")
)$optimality_status
```

## What if one landmark is an outlier?

Squared $L_2$ lets a single bad landmark pull the consensus. Huber
IRLS down-weights the landmark vector and reports a different claim:
still stationarity, never Ling's dual certificate.

```{r robust-fit}
contaminated <- recordings
contaminated$site_c[3, ] <- contaminated$site_c[3, ] + c(4, -3.5)
l2 <- gpa(contaminated, transform = proc_orthogonal("O"))
huber <- gpa(
  contaminated,
  transform = proc_orthogonal("O"),
  loss = proc_huber(k = 0.75)
)
aug <- augment.gpa_fit(huber)
weights_c <- aug$weight[aug$configuration == "site_c"]
data.frame(
  engine = c(l2$solver, huber$solver),
  claim = c(l2$optimality_status, huber$optimality_status),
  certificate = c(certify(l2)$status, certify(huber)$status),
  site_c_landmark3_weight = c(NA_real_, weights_c[[3L]])
)
```

```{r check-huber, include = FALSE}
stopifnot(identical(huber$solver, "irls"))
stopifnot(identical(huber$optimality_status, "blockwise_stationary"))
stopifnot(identical(certify(huber)$status, "unavailable"))
stopifnot(weights_c[[3L]] < min(weights_c[-3L]))
```

The influence plot shows the same weights the IRLS step used.

```{r plot-influence}
plot(huber, type = "influence")
```

## What if the recordings differ by more than similarity?

Affine, linear-basis warps, and thin-plate splines are a separate
engine. They solve a constrained eigenproblem in reference space.
`reference_covariance` is $\Lambda$ in $M^\top M=\Lambda$. It is a
modeling choice, never estimated invisibly. Success here is a global
eigen-solution **for that formulation**, not "GPA is globally solved."

```{r affine-engine}
aff <- gpa(recordings, transform = proc_affine(reference_covariance = 1))
c(
  solver = aff$solver,
  claim = aff$optimality_status,
  certificate = certify(aff)$status,
  lambda = paste(aff$reference_covariance, collapse = ",")
)
```

```{r check-affine, include = FALSE}
stopifnot(identical(aff$solver, "lbw_eigen"))
stopifnot(identical(aff$optimality_status, "exact_closed_form"))
stopifnot(!identical(certify(aff)$status, "certified_global"))
M <- consensus(aff)
stopifnot(isTRUE(all.equal(unname(diag(crossprod(M))), c(1, 1), tolerance = 1e-6)))
```

A TPS fit can fold. Affine $\det<0$ is orientation, not folding.
`datum_space_error()` exists only when the transform has an exact
inverse; TPS refuses that inverse.

```{r inverse-and-folding}
tps <- gpa(recordings, transform = proc_tps(smoothness = 1))
c(
  affine_datum = datum_space_error(aff),
  tps_folded = tps$folding$any_folded
)
```

```{r tps-inverse, error = TRUE}
inverse_proc_transform(transformations(tps)[["site_a"]])
```

## What should I reach for next?

- `?gpa` and `?explain_solver` for engine dispatch
- `?infer` for bootstrap or tangent standard errors after fitting
- `?cross_validate` and `?tune_smoothness` for held-out checks
- `?quotient_distance` when you need a gauge-invariant comparison
- the guarantee matrix in `docs/spec/01-solver-guarantee-matrix.md`
